Gradual Soundness: Lessons from Static Python

### 2021-12-21 pj-7.1/paper.tex
pj: quick note (in comment) about ctypes and dyn

Like most gradual languages, \SP{} includes a dynamic type
that allows untyped expressions within a statically-typed context.
Whenever an expression or variable lacks a type annotation,
\SP{} uses the dynamic type as a default and skips most compile-time checks.
Every untyped Python program is thus a well-formed \SP{} program.
Furthermore, programmers can omit the annotations for specific
function arguments without losing the benefits of other type checks.
Freedom to omit annotations helps programmers convert modules to \SP{}
with little initial refactoring.

In addition to serving as a default, the dynamic type also replaces any undefined types.
For example, suppose that the type \code{X} is undefined in \SP{} and that a
program contains the annotation \code{ChkDict[int, X]}, which describes a
dictionary with integer keys and \code{X}-typed values.
\SP{} interprets this annotation as \code{ChkDict[int, dynamic]} and allows any sort
of values inside the dictionary at runtime.
To an outsider, this behavior might seem odd: why not reject undefined types with an error?
The reason is to reduce friction with the Pyre annotations that are already
widely-used at Instagram.
Pyre is a mature static type system that can check more types than \SP{} knows how to enforce
efficiently, e.g., the type \code{Union[t1, ..., tn]} of arbitrary-width unions.
By replacing such types with dynamic, \SP{} can accept Pyre code that it does not yet fully understand.

The dynamic type is not, however, the flexible dynamic type provided
by \emph{true} gradual languages~\cite{svcb-snapl-2015}.
Replacing part of a type with dynamic can lead to both compile-time errors
and run-time errors.
In other words, \SP{} satisfies neither the static nor the dynamic gradual guarantee.\footnote{We
assume a standard type precision relation. Of course, one could argue that these programs
do not break the gradual guarantees if type precision does not allow our uses of the dynamic type.}
\Cref{fig:gg-failure} presents two examples, one for each kind of failure:
\begin{itemize}
  \item
    \Cref{f:gg-failure-stat} presents a fully-typed class and a partially-typed subclass.
    The subclass definition raises a compile-time error because it attempts to override
    the typed \code{m1} method with another method that returns the dynamic type.
  \item
    \Cref{f:gg-failure-dyn} presents a variant of \cref{fig:sp-example} that
    uses the type \code{ChkDict} for concrete-typed dictionaries (\cref{s:checked-type}).
    It sends a dictionary with integer values to a function that expects
    dictionaries with dynamic values.
    At runtime, the function rejects this argument because its type is not an
    exact match.
\end{itemize}
These behaviors are a consequence of \SP{} being driven by the needs of an
existing codebase.
The literature on gradual types has served only as a guide, not a mandate.

@@

\subsubsection{Tailored Types}

>>>>

Like most gradual languages, \SP{} includes a dynamic type
that allows untyped expressions within a statically-typed context.
Whenever an expression or variable lacks a type annotation, \SP{} uses the
dynamic type as a default and skips most (but not all!) compile-time checks.
Every untyped Python program is thus a well-formed \SP{} program.
Furthermore, programmers can omit the annotations for specific
function arguments without losing the benefits of other type checks.
Freedom to omit annotations helps programmers convert modules to \SP{}
with little initial refactoring.

In addition to serving as a default, the dynamic type also replaces any undefined types.
For example, suppose that the type \code{X} is undefined in \SP{} and that a
program contains the annotation \code{ChkDict[int, X]}, which describes a
dictionary with integer keys and \code{X}-typed values.
\SP{} interprets this annotation as \code{ChkDict[int, dynamic]} and allows any sort
of values inside the dictionary at runtime.

At first glance, this behavior might seem odd: why not reject undefined types with an error?
The reason is to reduce friction with the Pyre annotations that are already
widely-used at Instagram.
Pyre is a mature static type system that can check more types than \SP{} knows how to enforce
efficiently, e.g., the type \code{Union[t1, ..., tn]} of arbitrary-width unions~(\cref{s:idiomatic-types}).
By replacing such types with dynamic, \SP{} can accept Pyre code that it does not yet fully understand.

%% intro sentence
The dynamic type is also different than the flexible dynamic type provided
by true gradual languages~\cite{svcb-snapl-2015}.
Replacing part of a type with dynamic can lead to both compile-time errors
and run-time errors.
In other words, \SP{} satisfies neither the static nor the dynamic gradual guarantee
for a standard type precision relation.
%% \footnote{We
%% assume a standard type precision relation. Of course, one could argue that these programs
%% do not break the gradual guarantees for a type precision relation that is tailored to \SP{}.}
%%   %% tailored = describes exactly the migrations that SP allows
\Cref{fig:gg-failure} presents two examples, one for each kind of failure:
%% \footnote{In addition, C types~(\cref{s:ctypes}) are incompatible with type dynamic.}
\begin{itemize}
  \item
    \Cref{f:gg-failure-stat} presents a fully-typed class and a partially-typed subclass.
    The subclass definition raises a compile-time error because it attempts to override
    the typed \code{m1} method with another method that returns the dynamic type.
  \item
    \Cref{f:gg-failure-dyn} presents a variant of \cref{fig:sp-example} that
    uses the type \code{ChkDict} for concrete-typed dictionaries (\cref{s:checked-type}).
    It sends a dictionary with integer values to a function that expects
    dictionaries with dynamic values.
    At runtime, the function rejects this argument because its type is not an
    exact match.
\end{itemize}
These behaviors are a consequence of \SP{} being driven by the needs of an
existing codebase.
The literature on gradual types has served only as a guide, not a mandate.

@@

\subsubsection{Tailored Types}
\label{s:idiomatic-types}
